---
title: "Homework 4: Bayesian Inference for Mixture Models"
author: "BIOS 731 – Advanced Statistical Computing"
format:
  pdf:
    colorlinks: true
    urlcolor: blue
    include-in-header:
      - text: |
          \usepackage{multirow}
execute:
  echo: true
  warning: false
  message: false
---


```{r, echo = FALSE}
library(tidyverse)
```

## Context and learning objectives

This assignment reinforces ideas in the **Bayesian Computing** and **Cluster Computing** modules. We focus specifically on implementing a Bayesian K-Mixture Model using Gibb's sampling and variational inference.

## Due Date and Submission

Submit (via Canvas) a PDF knitted from a `.Rmd` or `.qmd` file. Your PDF should include the web address of the GitHub repository containing your work for this assignment. Due date is Wednesday, 4/8 at 10:00AM.

## Point distribution

```{r, echo = FALSE}
knitr::kable(
  data.frame(
    Problem = c("Problem 0", "Problem 1", "Problem 2", "Problem 3", "Problem 4"),
    Points = c(10, 20, 20, 20, 30)
  ),
  align = c("l", "r")
)
```

## Problem 0 

This "problem" focuses on the structure of your submission, especially the **use of git and GitHub** for reproducibility, **R Projects** to organize your work, **Quarto/R Markdown** to write reproducible reports, **relative paths** to load local files, and reasonable naming conventions for your files.

To that end:

- Create a public GitHub repository and a local R Project. I suggest naming the repo/directory `bios731_hw3_YourLastName` (e.g., `bios731_hw3_wrobel`).
- Push your **entire project folder** to GitHub.
- Submit a PDF knitted from your `.Rmd`/`.qmd` file to Canvas.
  - Your solutions should be implemented in your `.Rmd`/`.qmd` file.
  - Your git commit history should reflect your workflow (i.e., multiple meaningful commits; avoid a single “final” commit with all work).

## Bayesian K-Mixture Model


We consider a Bayesian Mixture Model  of unit-variance univariate Gaussians. There are:

- $K$ mixture components, corresponding to $K$ Gaussian distributions with means $\mathbf{\mu} = \{\mu_1,\ldots,\mu_K\}$
- cluster assignments $c_i \in \{1,\dots,n\}$ 
- observations $y_i \in \{1,\dots,n\}$


Data generating model is

\begin{align*}
\mu &\sim N(0, \sigma^2)\\
c_i &\sim Categorical(1/K. \ldots, 1/K)\\
y_i|c_i, \mu &\sim N(c_i^T\mu, 1)
\end{align*}




## Problem 1: Gibbs sampler derivations and implementation

Derive a Gibb's sampler for the posterior distribution, $p(\mu, \textbf{c}|y)$.
Show the derivations and resulting Gibb's sampler updates.  Specifically, derive:

- $p(c_i|y_i, \mu)$
- $p(\mu_k|y, c)$



Implement the Gibb's sampler in R. Your Gibb's R function should

- run the sampler
- store posterior draws
- return posterior summaries (estimates of the cluster assignments and means).


## Problem 2: Variational inference derivations

Assume a mean‑field approximation. Derive a CAVI algorithm for the posterior distribution, $p(\mu, \textbf{x}|y)$. Show explicit derivations and resulting updates. Implement the algorithm in R. You may refer to [Blei, 2017](https://www.tandfonline.com/doi/full/10.1080/01621459.2017.1285773) for details.

Your variational inference R function should

- monitor convergence using the ELBO
- return final estimates of the cluster assignments and means


## Problem 3: Fit to data

Implement the Gibb's sampler and variational inference updates in R.  Fit to the Old Faithful waiting times dataset used in the EM lecture.

Compare computation times and estimates from variational bayes to mean parameter estimates from the Gibb's sampler.  For Gibb's sampler, also do diagnostic plots, fit multiple chains, comment on convergence and effective sample size.


## Problem 4: Perform a simulation study on the cluster



In our simulation, we want to 

* Generate data from a 4-part Gaussian mixture model with true means $\mu = \{0, 5, 10, 20\}$ and unit variance.
  * Evaluate estimation of $\mu$ through bias, coverage, and computation time
  * We will compare the Gibbs sampler and Variational Bayes algorithms defined above
  

* Evaluate these properties at sample sizes $n =\{100, 1000, 10000\}$
  - Use Gibbs chains of length 10000 and discard the first 2000 samples as burn in
  - Use nsim = 500
  

We will execute this simulation study on the cluster. For full credit, make sure to implement the following:

**Workflow:**
* Use structured scripts and subfolders following guidance from the cluster computing project organization lecture
* Instead of parallelizing your simulation scenarios (as in HW1), each simulation scenario should be assigned a different JOBID on the cluster.


**Presenting results:**

Create plots with *Monte Carlo standard error bars* to summarize the following:

- Bias of $\hat{\mu}$
- Coverage of $\hat{\mu}$


Write 1-2 paragraphs summarizing these results.

